To be FAIR: Theory Specification Needs an Update

### 2025-06-07 manuscript.Rmd
update 2025-06-07 10:38:20.297695

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Conceptualization
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Maximilian S. Ernst"
    affiliation: "2,4"
    role:
      - Writing – review & editing
      - Investigation
  - name: "Noah N.N. van Dongen"
    affiliation: "5"
    role:
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "6"
    role:
      - Conceptualization
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,7"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University, dept. Methodology & Statistics"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "5"
    institution   : "University of Amsterdam, the Netherlands"
  - id            : "6"
    institution   : "Ludwig-Maximilians-Universität München, Germany"
  - id            : "7"
    institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
authornote: |
  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.

@@

With regard to the state of Interoperability in psychology, 
Kurt Lewin's [-@lewinPsychologyProcessGroup1943] adage "there's nothing as practical as a good theory" paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
But, as we argued, this is not the case.
The examples of X-Interoperability offered in Table \@ref(tab:tabmeehl) illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-Interoperable within software used for analysis.
For example, Interoperable theory could be used
to select control variables for causal inference [@cinelliCrashCourseGood2022],
or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data [@peikertReproducibleResearchTutorial2021],
or to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021] which could be automatically evaluated through integration testing [@vanlissaUsingEndpointsCheck2023].
Furthermore, theories can be X-Interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
For example, it should be possible to embed a theory about emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].

@@

One issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":
it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [@bringmannBackBasicsImportance2022].
The "jingle-jangle" fallacy is common in psychology:
the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
FAIR theory can help address the measurement crisis:
since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of measurement.

>>>>

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Conceptualization
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Maximilian S. Ernst"
    affiliation: "2,4"
    role:
      - Writing – review & editing
      - Investigation
  - name: "Noah N.N. van Dongen"
    affiliation: "5"
    role:
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "6"
    role:
      - Conceptualization
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,7"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University department of Methodology & Statistics, Tilburg, The Netherlands"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "5"
    institution   : "University of Amsterdam, Amsterdam, The Netherlands"
  - id            : "6"
    institution   : "Ludwig-Maximilians-Universität München, München, Germany"
  - id            : "7"
    institution   : "MSB Medical School Berlin Department of Psychology, Berlin, Germany"
authornote: |
  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.

@@

With regard to the state of Interoperability in psychology, 
Kurt Lewin's [-@lewinPsychologyProcessGroup1943] adage "there's nothing as practical as a good theory" paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
But, as we argued, contemporary practice falls short of this ideal.
The examples of X-Interoperability offered in Table \@ref(tab:tabmeehl) illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-Interoperable within software used for analysis.
For example, Interoperable theory could be used
to select control variables for causal inference [@cinelliCrashCourseGood2022],
or to preregister a study with an explicit derivation chain from theory to hypothesis,
as well as an inferential procedure that would suggest specific modifications to theory after analyzing empirical data [@peikertReproducibleResearchTutorial2021],
or to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021] which could be automatically evaluated through integration testing [@vanlissaUsingEndpointsCheck2023].
Furthermore, theories can be X-Interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
For example, it should be possible to embed a theory about emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].

@@

<span id="fairtheoryprereg">One important future direction is embedding FAIR theories withing existing open science methodologies. 
For example, consider how FAIR theory relates to preregistration.
These practices are clearly distinct but complimentary.
The purpose of FAIR theory is to communicate general principles and expectations about a given phenomenon,
and to provide infrastructure for explicitly deriving hypotheses from specific theories and revising those theories in light of empirical results.
The purpose of preregistration, by contrast, is to eliminate inductive bias from hypothesis tests and increase trust in their outcomes [@peikertWhyDoesPreregistration2023].
These practices differ in their level of abstraction:
FAIR theories cut across studies on a given phenomenon,
whereas preregistrations are specific implementations of hypothesis tests,
with a specific study design, analysis plan, and - optionally - a fully reproducible analysis pipeline. 
These practices complement each other:
authors can make the derivation chain from theory to hypothesis more explicit by citing a specific FAIR theory in their preregistration.
Moreover, it is possible to preregister an inferential procedure that would require revising the theory after observing data,
or even to have proponents and detractors of a theory review a registered report of such a test.
In short, combining FAIR theory with preregistration and other existing open science practices has the potential to strengthen the epistemic cycle of prediction, testing, and revision, moving us closer to a cumulative science.</span>

Another future direction is the intersection between the aforementioned "theory crisis" and the related "measurement crisis" pertaining to the lack of clarity, consistency, and validity in the operationalization of theoretical constructs [@bringmannBackBasicsImportance2022].
Since FAIR theories can reference other theories and resources,
it is possible to attach references to specific measurement instruments (or even theories of measurement) to constructs named in a theory.
<span id="jinglejangle">FAIR theories can also help address "jingle- and jangle fallacies" in psychology,
which are ambiguities that arise from using the same term for different constructs, or conversely, using different terms for the same construct [@songLiteratureReviewingAddressing2021].
By explicitly referencing operational definitions in FAIR theories,
such jingle-jangle fallacies would come to light and could ultimately be resolved.</span>
